Activity Monitoring
========================
load required packages:
```{r, echo=FALSE,results='hide',warning=FALSE}
required <- c('ggplot2','knitr')
lapply(required,require,character.only=TRUE)
```

```{r setoptions, echo=FALSE, results='hide'}
opts_chunk$set(cache=TRUE,warning=FALSE,message=FALSE)
```

Download and read data into R:
```{r}
fileurl <- 'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
if(!file.exists('mon.zip')) {download.file(fileurl, destfile = 'mon.zip')}
downdate <- date()
data <- read.csv(unz('mon.zip','activity.csv'))
```
Data was downloaded from link above on `r downdate`.

##Data Cleaning
```{r}
str(data) #look at data
data$date <- as.Date(as.character(data$date),format='%Y-%m-%d') #convert to Date format
```

##Data Analysis
###Part1
What is mean total number of steps taken per day?  
* Historgram for total number of steps per day  
* Mean and median of total number of steps per day 
```{r}
tsum<- tapply(data$steps,data$date,sum,na.rm=T)
dfsum<-data.frame(tsum)
p1<- ggplot(data=dfsum,aes(x=tsum))
p1<- p1+geom_histogram()
p1<- p1+labs(x='Total Number of Steps',y='Frequency')
p1
outmat <- matrix(nrow=length(unique(data$date)),ncol=2,dimnames = list(as.character(unique(data$date)),c('mean','median')))
mm <- sapply(list('mean','median'),function(x){ tmp<-tapply(data$steps, data$date, x,na.rm=T); outmat[,x]<<- tmp})
outmat #mean and median per day
mean(dfsum$tsum) #mean of total number of steps for the entire dataset
median(dfsum$tsum) #median of total number of steps for the entire dataset
```

###part2
What is the average daily activity pattern?  
* time series plot (interval vs mean number of steps)  
* which time interval contains the maximum number of steps  
```{r}
t<-tapply(data$steps,data$interval,mean,na.rm=T)
t[which.max(t)]
df <-data.frame(t) #converting t to dataframe for plotting purposes
df$inv <- as.numeric(rownames(df))
p2 <-ggplot(data=df,aes(x=inv,y=t))
p2<- p2+geom_line()
p2<- p2+labs(x='Interval',y='average number of steps')
p2
```

###Part3
* total number of missing values  
* impute the missing data  
* redo Part1 with the new data set  
* what is the impact of data filling?  

```{r}
sum(is.na(data)) #total number of missing data
sapply(data, function(x) sum(is.na(x))) #by variable
```
All the missing data is in the first column, steps

Impute the missing values with the average number of steps for the entire dataframe
```{r}
imp <- mean(data$steps,na.rm=T)
data1 <- data #copy the original data
data1[is.na(data),'steps'] <- imp
```
Double check to see if there are any missing data left:
```{r}
sum(is.na(data1))
```

Histogram:
```{r}
tsum1<- tapply(data1$steps,data1$date,sum,na.rm=T)
dfsum1<-data.frame(tsum1)
p3<- ggplot(data=dfsum1,aes(x=tsum1))
p3<- p3+geom_histogram()
p3<- p3+labs(x='Total Number of Steps',y='Frequency')
p3
```

the peak of the plot corresponds to the mean of the dataset because all the missing values were replaced with the mean of the dataset.


Mean and median:
```{r}
outmat1 <- matrix(nrow=length(unique(data1$date)),ncol=2,dimnames = list(as.character(unique(data1$date)),c('mean','median')))
mm1 <- sapply(list('mean','median'),function(x){ tmp<-tapply(data1$steps, data1$date, x,na.rm=T); outmat1[,x]<<- tmp})
outmat1 #mean and median per day
mean(dfsum1$tsum1) #mean of total number of steps for the entire dataset
median(dfsum1$tsum1) #median of total number of steps for the entire dataset
```

let's compare the summaries of the mean:
```{r}
summary(outmat[,'mean'])
summary(outmat1[,'mean'])
```
As expected, the figures didn't change much since a simple imputation was done to fill in the missing data with the average number of total steps for the entire dataframe.

###Part4
Are there differences in activity patterns between weekdays and weekends?  
* create a new variable for weekdays  
* make a panel plot  

```{r}
data1$wday <- weekdays(data1$date)
data1$wday <- ifelse(data1$wday %in% c('Saturday','Sunday'),'Weekend','Weekday')
t1<-aggregate(steps~interval+wday,data=data1,mean,na.rm=T)
p4 <- ggplot(data=t1,aes(x=interval,y=steps))
p4 <- p4+geom_line()+facet_grid(wday~.)
p4
```
